Lowestcase and uppestcase letters: Advances in derp learning

### 2021-03-31 trunk/lowercase/paper/lowercase.tex
Various fixes / tools for video

\title{Lowestcase and Uppestcase letters: Advances in Derp Learning}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
Appears in \upsigbovik~2021 with the
OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 em}

@@

\paragraph{Black-box optimization.} If you have a function and want
to find arguments that minimize its output, the most efficient
techniques are generally those like gradient descent. (In fact, the
backpropagation algorithm we use to train the neural network in
Section~\ref{sec:neural} is gradient descent on the function that
takes the model weights and produces an error value for each output
node.) The problem with this is that you need to do some math to
compute the derivative of the function, and anyway you need to deal
with fiddly bits (Section~\ref{sec:fiddly}) unless the function is
convex and smooth, which it will not be. If you don't want to deal
with that, and have a fast computer (and who doesn't?), black-box
optimization algorithms are worth considering. Here, the
interface\footnote{ Here a simplfied wrapper around
  BiteOpt~\cite{biteopt} in my {\tt cc-lib} library. See
  \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/cc-lib/opt/}.}
is just something like (C++):

@@

As this was training, I worked on another improvement. Ideally we
would compute the difference between the predicted shape and the
expected shape, regardless of how they're drawn. Aside from being a
bit computationally challenging, this won't really work because we
need to attribute error directly to each output in order to actually
update the model in training. I spent a valuable vacation day writing
a routine to compute the best alignment of points between the actual
and expected outputs (Figure~\ref{fig:trainingrotate}). Aside from being
harder than it looked, my alignment code ended up being pretty slow
relative to the rest of training, even worse since it ran on the CPU
instead of GPU, which reduced the training speed by $50\%$. I let it
run for 80,000 rounds, some 331 million training examples, but
eventually got bored of waiting on this approach that was slow to
train and seemed like a complicated version of an bad, oversimplified
approach. So, I control-C'd that thing and threw this whole endeavour
in the trash! But I must have confused the Recycle~Bin icon with the
fairly complicated export-to-TrueType Font process that I built,
because I ran the model on the venerable Futura~\cite{futura} font and
generated {\bf Futurda} (Figure~\ref{fig:futurda}).
%
% \begin{figure}[ht]
% \centering
%   \includegraphics[width=0.5 \linewidth]{rotatej}
% \caption{ Example alignment (from debug UI) of predicted shape (blue)
%   to expected shape (green). We require each point to be mapped (red)
%   to a point from the expected contour in a monotonic order (but
%   several can be mapped to the same one), so that we can attribute
%   error to each point. I hope it was fun to write this code because
%   it was too slow to do pretty much anything but generate this figure!
% } \label{fig:rotatej}
% \end{figure}
%

@@

I also adapted my weird home-grown software to train the
\makeuppercase\ and \makelowercase\ models simultanously. Two models
fit easily in GPU memory, with plenty of space for a stream of
training data (one training instance is only about 10kb). The only
challenging thing is arranging for them to feed each other generated
``inversion'' examples (Figure~\ref{fig:cotraining}), but this is just
a matter of programming, thank god. I should remember to do projects
that are mostly a matter of programming. Each round, $25\%$ of the
batch consists of inverted examples from the symmetric model's output
from a recent round. Training happens asynchronously, but I make sure
that one model is not allowed to get more than 2 rounds ahead of the
other, because I want this feedback loop to be somewhat tight.

@@

\paragraph{Generating features.} On the other side, randomly sampling
pixels from the input SDF does work, but I superstitiously believed
that it might be better to have more spatially meaningful features. I
wrote a program to generate a bunch of random simple features (one
line/blob with positive weights, one line/blob with negative weights).
It then chooses a set of them that are both {\em good} (maximum
standard deviation on a sample of training data) and {\em different
  from one another} (redundant or even partially redundant features
are less valuable). It was nice to satisfy my supersition, and the
dark wizard of superstitious fiddling with neural networks in the
hope that they do the thing was pleased as well. The features are
at least handsome (Figure~\ref{fig:makefeatures}). Creativity is
enriching!

@@

\begin{figure}[ht!b]
\centering
  \includegraphics[width=0.99 \linewidth]{lastlayer}
  \caption{ The bottom-right corner of the weight matrix for the final
    layer of the network. This layer was added to the network after
    1.36 million rounds, initially as the identity matrix, and so can
    be thought of partly as a correction of the network's output prior
    to that round (though the remainder of the network continues to
    evolve). The x-axis is the output nodes, and the y-axis is the
    nodes of the previous layer. Note for example that the last 26
    columns look pretty different; these are the predictors for the 26
    letters, which occur in the output after the SDF pixels. Green
    means positive and red means negative, so if you are looking at
    this in a black-and-white printout, that may explain your current
    confusion. The exact diagonal is a strong green, close to 1.0,
    although over time these weights do diverge from the identity
    somewhat. In the bottom-right corner of size $26^2$, we are
    looking at how the 26 letter predictors are derived from the
    previous layer's predictions. We see that most letters are
    negatively correlated (makes sense; only one will ever be 1.0)
    although there are some oddities (probably because it found some
    other, better correlates). Nodes on this new layer have dense
    references to these 26 predictions on the previous; this means
    that the bottom row kind of represents biases for each of the 26
    letters (what does an average 'e' look like?). I also included a
    dense region above that, but this appears to have simply evolved
    the same way as other $36^2$ chunks have (the rest are sparse).
    These chunks have a large amount of spatial similarity (suggesting
    that the sparse sampling would be adequate), with a meaning like
    ``if this area of the image is bright, then this pixel should
    be less bright.'' It is interesting that the pixels immediately
    next to the diagonal are almost always strongly negative
    (sharpening operation). ``Thank you for attending my TED talk.''
    ---\,Figure~\ref{fig:lastlayer}}
  \label{fig:lastlayer}
\end{figure}

@@

Now that we have these expensive models, we can use them to make
arbitrary letterforms uppercase or lowercase. The output is readily
rasterized (using the standard thresholding approach for SDFs) but
we'd actually like to have vector representations so that we can
use them as TrueType fonts for our desktop publishing needs.

@@

This begs us to ask the question: What shapes {\em do} look like
letters? Since the models will tell us, I can just search over shapes
and ask them. The first thing I tried was to just generate random
shapes and optimize their parameters to produce the highest possible
prediction for the target letter, and the lowest possible prediction
for the rest. This preduced results that are fully bonkers
(Figure~\ref{fig:bonkersf}).

@@

% figure? drop it?

@@

People often stop me on the street to ask, Tom, Why did you spend so
much time and energy on this useless SIGBOVIK project? To which I say,
Ha! At least I am not wasting my time {\it reading} SIGBOVIK papers or
talking to strangers on the street! and run off. Although the main
purpose of SIGBOVIK is to confound bibliometrics with ambiguously
good-faith and high-quality research published in a clearly satirical
but superfically decorous venue associated with a traditionally
esteemed university, it is also possible for such work to have
practical applications in arts and spycraft. You simply need to give
it some thought.

For example, it is well known to {\sf internet troIIs} and {\sf DOMAlN
  NAME PHlSHERS} that the uppercase \letterform{i} and loweracse
\letterform{L} are indistinguishable in many fonts, allowing for
various Tomfoolery.\footnote{I actually wrote ``troiis'' and ``domaln
  name phlshers'', hehe!} This can also be used for
steganography---hiding messages inside text without the use of em
dashes---by selectively replacing letters with their alternates. Each
replacement only encodes one bit of information, however. With the
generic ability to uppercase and lowercase letterforms, we can exploit
this ambiguity to generate a large variety of letterforms that can be
used like this.

@@

Iterating the \makeuppercase\ model with 32-bit floats does not form
any cycles in 25,000,000 iterations, nor does it if the intermediates
are quantized to 8-bit ints. This is curious because under visual
observation the sequence does appear periodic, and in fact seems to be
the same characteristic loop reached by all letters. Presumably there
are some oscillations with long, relatively prime periods or even some
pixels that are monotonically growing or shrinking, but very slowly.
However, if I render the SDF as a 1-bit bitmap at $2\times$ scale, I
get a perfect cycle of 132 frames, which appears to be two passes
through the main characteristic loop, but slightly different each
time. This is not a strange property for a letter to have; for example
a typical \letterform{B} has two copies of the same basic idea in it.

.... XXX more here ...

% lowercase uppercase letters etc.
% iterate this a lot of times?
% "to infinity, but let's stop there"

% run it to infinity to see the uppestcase, etc.
% if these vanish, then we have a fixed point.
% if not, or perhaps by doing some auto-gamma, then
% it's like a fractal zoomer, and we can set the
% record on the "most uppercase letter" by running
% a jazillion times.

% lowestcase is just dust, not surprising
% uppestcase is actually a fixed point (not FixedSys)
% ... so it's like a 3D manifold

% bristol stool chart

\section{Conclusion}
... XXX ...
This fixed point
suggests that the lowestcase letter is already on your keyboard,
but it is not a point, it is just \letterform{:}.

\paragraph{Futura Work}
I think I pretty much beat this one to death, honestly.

\paragraph{Acknowledgements}
For this project I used \verb+stb_truetype.h+ from the excellent stb
collection of single-file libraries~\cite{stb}. I did push its limits
somewhat and ``automatically discovered'' assertion failures and other
crashes (e.g.~during blackbox optimization on all 100k fonts), but it
saved a lot of time. This library only helps with reading the fonts
and generating SDFs. To generate TTFs, I had to write my own pipeline,
which generates FontForge's .SFD (a typo factory right there) files,
and then did the final export with FontForge~\cite{fontforge}. Thanks
Jason Reed for this suggestion, without which I would probably still
be developing my own TTF file writer and custom hinting engine. Finally,
I would like to thank the pseudonymous SIGBOVIK program committee for
overseeing the proceedings, and the \upsigbovik\ program committee
committee for overseeing {\em them}.

>>>>

\title{Lowestcase and uppestcase letters: Advances in derp learning}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
Appears in \upsigbovik~2021 with the
OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 em}

@@

\paragraph{Black-box optimization.} If you have a function and want
to find arguments that minimize its output, the most efficient
techniques are generally those like gradient descent. (In fact, the
backpropagation algorithm we use to train the neural network in
Section~\ref{sec:neural} is gradient descent on the function that
takes the model weights and produces an error value for each output
node.) The problem with this is that you need to do some math to
compute the derivative of the function, and anyway you need to deal
with fiddly bits (Section~\ref{sec:fiddly}) unless the function is
convex and smooth, which it will not be. If you don't want to deal
with that, and have a fast computer (and who doesn't?), black-box
optimization algorithms are worth considering. Here, the
interface\footnote{ Here a simplified wrapper around
  BiteOpt~\cite{biteopt} in my {\tt cc-lib} library. See
  \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/cc-lib/opt/}.}
is just something like (C++):

@@

As this was training, I worked on another improvement. Ideally we
would compute the difference between the predicted shape and the
expected shape, regardless of how they're drawn. Aside from being a
bit computationally challenging, this won't really work because we
need to attribute error directly to each output in order to actually
update the model in training. I spent a valuable vacation day writing
a routine to compute the best alignment of points between the actual
and expected outputs (Figure~\ref{fig:trainingrotate}). Aside from being
harder than it looked, my alignment code ended up being pretty slow
relative to the rest of training, even worse since it ran on the CPU
instead of GPU, which reduced the training speed by $50\%$. I let it
run for 80,000 rounds, some 331 million training examples, but
eventually got bored of waiting on this approach that was slow to
train and seemed like a complicated version of an bad, oversimplified
approach. So, I control-C'd that thing and threw this whole endeavor
in the trash! But I must have confused the Recycle~Bin icon with the
fairly complicated export-to-TrueType Font process that I built,
because I ran the model on the venerable Futura~\cite{futura} font and
generated {\bf Futurda} (Figure~\ref{fig:futurda}).

@@

I also adapted my weird home-grown software to train the
\makeuppercase\ and \makelowercase\ models simultaneously. Two models
fit easily in GPU memory, with plenty of space for a stream of
training data (one training instance is only about 10kb). The only
challenging thing is arranging for them to feed each other generated
``inversion'' examples (Figure~\ref{fig:cotraining}), but this is just
a matter of programming, thank god. I should remember to do projects
that are mostly a matter of programming. Each round, $25\%$ of the
batch consists of inverted examples from the symmetric model's output
from a recent round. Training happens asynchronously, but I make sure
that one model is not allowed to get more than 2 rounds ahead of the
other, because I want this feedback loop to be somewhat tight.

@@

\paragraph{Generating features.} On the other side, randomly sampling
pixels from the input SDF does work, but I superstitiously believed
that it might be better to have more spatially meaningful features. I
wrote a program to generate a bunch of random simple features (one
line/blob with positive weights, one line/blob with negative weights).
It then chooses a set of them that are both {\em good} (maximum
standard deviation on a sample of training data) and {\em different
  from one another} (redundant or even partially redundant features
are less valuable). It was nice to satisfy my superstition, and the
dark wizard of superstitious fiddling with neural networks in the
hope that they do the thing was pleased as well. The features are
at least handsome (Figure~\ref{fig:makefeatures}). Creativity is
enriching!

@@

\begin{figure}[ht!b]
\centering
  \includegraphics[width=0.99 \linewidth]{lastlayer}
  \caption{ The bottom-right corner of the weight matrix for the final
    layer of the network. This layer was added to the network after
    1.36 million rounds, initially as the identity matrix, and so can
    be thought of partly as a correction of the network's output prior
    to that round (though the remainder of the network continues to
    evolve). The x-axis is the output nodes, and the y-axis is the
    nodes of the previous layer. Note for example that the last 26
    columns look pretty different; these are the predictors for the 26
    letters, which occur in the output after the SDF pixels. Green
    means positive and red means negative, so if you are looking at
    this in a black-and-white printout, that may explain your current
    confusion. The exact diagonal is a strong green, close to 1.0,
    although over time these weights do diverge from the identity
    somewhat. In the bottom-right corner of size $26^2$, we are
    looking at how the 26 letter predictors are derived from the
    previous layer's predictions. We see that most letters are
    negatively correlated (makes sense; only one will ever be 1.0)
    although there are some oddities (probably because it found some
    other, better correlates). All nodes on this new layer have dense
    references to these 26 predictions on the previous; this means
    that the bottom 26 rows kind of represent biases for each of the 26
    letters (what does an average 'e' look like?). I also included a
    dense region above that, but this appears to have simply evolved
    the same way as other $36^2$ chunks have (the rest are sparse).
    These chunks have a large amount of spatial similarity (suggesting
    that the sparse sampling would be adequate), with a meaning like
    ``if this area of the image is bright, then this pixel should
    be less bright.'' It is interesting that the pixels immediately
    next to the diagonal are almost always strongly negative
    (sharpening operation). ``Thank you for attending my TED talk.''
    ---\,Figure~\ref{fig:lastlayer}}
  \label{fig:lastlayer}
\end{figure}

@@

Now that we have these expensive models, we can use them to make
arbitrary letterforms uppercase or lowercase. The output is readily
rasterized (using the standard thresholding approach for SDFs) but
we'd actually like to have vector representations so that we can
download the TTF files and clog up our fonts menu forever.

@@

This begs us to ask the question: What shapes {\em do} look like
letters? Since the models will tell us, I can just search over shapes
and ask them. The first thing I tried was to just generate random
shapes and optimize their parameters to produce the highest possible
prediction for the target letter, and the lowest possible prediction
for the rest. This produced results that are fully bonkers
(Figure~\ref{fig:bonkersf}).

@@

@@

People often stop me on the street to ask, Tom, Why did you spend so
much time and energy on this useless SIGBOVIK project? To which I say,
Ha! At least I am not wasting my time {\it reading} SIGBOVIK papers or
talking to strangers on the street! and run off. Although the main
purpose of SIGBOVIK is to confound bibliometrics with ambiguously
good-faith and high-quality research published in a clearly satirical
but superficially decorous venue associated with a traditionally
esteemed university, it is also possible for such work to have
practical applications in arts and spycraft. You simply need to give
it some thought.

For example, it is well known to {\sf internet troIIs} and {\sf DOMAlN
  NAME PHlSHERS} that the uppercase \letterform{i} and lowercase
\letterform{L} are indistinguishable in many fonts, allowing for
various Tomfoolery.\footnote{I actually wrote ``troiis'' and ``domaln
  name phlshers'', hehe!} This can also be used for
steganography---hiding messages inside text without the use of em
dashes---by selectively replacing letters with their alternates. Each
replacement only encodes one bit of information, however. With the
generic ability to uppercase and lowercase letterforms, we can exploit
this ambiguity to generate a large variety of letterforms that can be
used like this.

@@

Iterating the \makeuppercase\ model with 32-bit floats does not form
any cycles in 25,000,000 iterations, nor does it if the intermediates
are quantized to 8-bit ints. This is curious because under visual
observation the sequence does appear periodic, and in fact seems to be
the same characteristic loop reached by all letters. Presumably there
are some oscillations with long, relatively prime periods or even some
pixels that are monotonically growing or shrinking, but very slowly.
However, if I render the SDF as a 1-bit bitmap at $2\times$ scale, I
get a perfect cycle of 132 frames. This appears to be two passes
through the main characteristic loop, but slightly different each
time. This is not a strange property for a letter to have; for example
a typical \letterform{B} has two copies of the same basic idea in it.

None of these alone could be considered the uppestcase letter, but
perhaps together they are? A natural way to include them all in one
shape is to stack them in 3D, as if each iteration is an MRI slice of
the brachial plexus.

To generate a 3D mesh, I stack the 132 SDFs in the z direction, and
this naturally yields a three-dimensional signed distance field
(trilinear interpolation). The classic ``marching cubes''
algorithm~\cite{lorensen1987marching} for mesh generation works
natively on such a field, and come to think of it, I probably could
have used that in two dimensions to trace the SDFs to vectors. Oh,
well. Fortunately there is enough spatial similarity between the
slices that it makes a reasonable 3D shape even without interpolation,
but sampled at a decent resolution and then cleaned up, it looks quite
nice; much better than the lowercase colon. Speaking of colons, it
rates approximately a 2.5 on the Bristol stool
scale~\cite{lewis1997bristol}. A 2D projection of the manifold is in
Figure~\ref{fig:uppestcase3d}. I 3D-printed it and am currently
working on a way of embedding it as an unusually tall key on my
keyboard that I can press whenever I wanted to express ULTIMATE ANGER.

\section{Conclusion}
We performed an exhaustive case analysis, exploring cases both more
upper- and lower- than ever been seen before. Sideways case was not
considered, as that is nonsense. We had modest success generating
upperercase and lowerercase letterforms through two neural network
models simultaneously trained to be each other's inverse, although
frankly it was much faster and more aesthetic to just do it by hand.
After really following through on the downloadables and taking some
needless excursions for effect, we saw that these models have
limits (at least informally). The lowestcase letter is already
on your keyboard; it is the ASCII eyeballs character \letterform{:}.
The uppestcase letter is not so easily typed, and perhaps that
is for the best.

\paragraph{Futura Work.}
I think I pretty much beat this one to death, honestly.

\paragraph{Acknowledgements.}
For this project I used \verb+stb_truetype.h+ from the excellent stb
collection of single-file libraries~\cite{stb}. I did push its limits
somewhat and ``automatically discovered'' assertion failures and other
crashes (e.g.~during blackbox optimization on all 100k fonts), but it
saved a lot of time. This library only helps with reading the fonts
and generating SDFs. To generate TTFs, I had to write my own pipeline,
which generates FontForge's .SFD (a typo factory right there) files,
and then did the final export with FontForge~\cite{fontforge}. Thanks
Jason Reed for this suggestion, without which I would probably still
be developing my own TTF file writer and custom hinting engine. Finally,
I would like to thank the pseudonymous SIGBOVIK program committee for
overseeing the proceedings, and the \upsigbovik\ program committee
committee for overseeing {\em them}.
